Effect of < v <= in percentile definition

### 2020-08-05 analysis/reports/less_than_equal.Rmd
working on ltet and add nb of unique skew and even values to pull_di outputs

---
title: "Figures and results for main manuscript"
output: 
    github_document:
       df_print: kable
---

@@

all_di <- read.csv(here::here("analysis", "reports", "all_di.csv"), stringsAsFactors = F)

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples),
         log_s0 = log(s0),
         log_n0 = log(n0)) %>%
  filter(n0 > s0,
         !singletons,
         dat %in% c("bbs", "fia_short", "fia_small", "gentry", "mcdb", "misc_abund_short")) %>%
  mutate(dat = ifelse(grepl(dat, pattern = "fia"), "fia", dat),
         dat = ifelse(dat == "misc_abund_short", "misc_abund", dat))
```

# Final dataset in S and N space

```{r final dataset s and n space}

ggplot(all_di, aes(x = log_s0, y = log_n0, color = dat)) +
  geom_point(alpha = .01) +
  geom_point(data = filter(all_di, log_nparts > log(50)), alpha = .4) +
  theme_bw() +
  scale_color_viridis_d() +
  ggtitle("Communities by S and N")

@@

```{r skew change}

ggplot(filter(all_di, s0 > 2), aes(x = skew_percentile_excl, y = skew_percentile, color = nsamples)) +
  geom_point() +
  facet_wrap(vars(dat))

all_di <- all_di  %>%
  mutate(skew_change = skew_percentile - skew_percentile_excl,
         even_change = simpson_percentile -
           simpson_percentile_excl)

ggplot(filter(all_di, s0 > 2), aes(x = log_nparts, y = skew_change)) +
  geom_point() +
  xlim(0,15) +
  facet_wrap(vars(dat))

@@

```{r hists}

>>>>

---
title: "Effect of < v <= in percentile definition"
author: Renata Diaz
date: "`r Sys.Date()`"
output: 
    github_document:
       df_print: kable
---

@@

@@

You can define the %ile score as either

- the % of values in a comparison vector that are **less than** a focal value
- the % of values that are **less than or equal to** that value

You see both definitions out in the world.

When I initially coded up the analysis, I used **<**. I didn't think about it much and as I was writing the manuscript I wrote **<=**. So that got me curious if it matters.

It *should not* matter too much as long as there are many unique values in the comparison vector. However, if there are only a handful, and the focal value is equal to an element of the comparison vector, using <= will cause the score to jump up. The jumps will be in increments of 100/n, with n the number of elements in the comparison vector. Meaning if there is only one value, < will give a score of 0 and <= will give a score of 100! And if there are ties, and the focal value is equal to the tie value, the jump will be (100 / n) * m, with m the number of elements in the comparison vector equal to the tie value.

#### Demo with toy data

```{r a handful of values, echo = T}

percentile_lt <- function(focal, comparison) {
  100 * (sum(comparison < focal) / length(comparison))
}

percentile_ltet <- function(focal, comparison) {
  100 * (sum(comparison <= focal) / length(comparison))
}

# Many values

many <- seq(1, 100, by = .1)

focal <- runif(1, 1, 100)
focal <- round(focal, 1) # rounding so focal == a value in many

percentile_lt(focal, many)
percentile_ltet(focal, many)

# A few values
few <- seq(1, 101, by = 20)

focal <- ceiling(focal / 20) * 20 + 1 # again, rounding so focal == a value in few

percentile_lt(focal, few)

percentile_ltet(focal, few)

# A few values with ties
few_with_ties <- c(seq(1, 81, by = 20), 41)

focal <- 41 # if focal == tie value, you get the largest possible jump

percentile_lt(focal, few_with_ties)

percentile_ltet(focal, few_with_ties)

# Just one value

one <- runif(1, 1, 100)

focal <- one # if there is only 1 element in one, focal MUST be that element

percentile_lt(focal, one)
percentile_ltet(focal, one)

@@

#### For our purposes

We can expect using <= to give us, if anything, higher percentile scores. We can expect this effect to be more pronounced for distributions with relatively few unique values. This is probably strongly correlated with the number of unique elements in the feasible set and the number of unique samples we found from the feasible set, but is not necessarily identical to it because two unique samples can have the same values for skewness or evenness. This will create **ties**, which can cause quite large jumps.

If the comparison vector is short, and/or has repeated values, we may get noticeably higher scores using <= than using <. Usually I think being short has a larger effect than having ties - unless you have a very long vector that is all ties, which doesn't happen in this context - and being short makes ties more impactful.

Especially if the feasible set is small, the focal value is guaranteed to be an element of the comparison vector. That is because the comparison vector, for small FS, is an exhaustive account of all possible values for the focal value.

<!-- ```{r real data, include=FALSE} -->

<!-- all_di <- read.csv(here::here("analysis", "reports", "all_di.csv"), stringsAsFactors = F) -->

<!-- all_di <- all_di %>% -->
<!--   mutate(log_nparts = log(gmp:::as.double.bigz(nparts)), -->
<!--          log_nsamples = log(nsamples), -->
<!--          log_s0 = log(s0), -->
<!--          log_n0 = log(n0)) %>% -->
<!--   filter(n0 > s0, -->
<!--          !singletons, -->
<!--          dat %in% c("bbs", "fia_short", "fia_small", "gentry", "mcdb", "misc_abund_short")) %>% -->
<!--   mutate(dat = ifelse(grepl(dat, pattern = "fia"), "fia", dat), -->
<!--          dat = ifelse(dat == "misc_abund_short", "misc_abund", dat)) -->
<!-- ``` -->

<!-- # Final dataset in S and N space -->

<!-- ```{r final dataset s and n space} -->

<!-- ggplot(all_di, aes(x = log_s0, y = log_n0, color = dat)) + -->
<!--   geom_point(alpha = .01) + -->
<!--   geom_point(data = filter(all_di, log_nparts > log(50)), alpha = .4) + -->
<!--   theme_bw() + -->
<!--   scale_color_viridis_d() + -->
<!--   ggtitle("Communities by S and N") -->

<!-- ``` -->

<!-- ```{r skew change} -->

<!-- ggplot(filter(all_di, s0 > 2), aes(x = skew_percentile_excl, y = skew_percentile, color = nsamples)) + -->
<!--   geom_point() + -->
<!--   facet_wrap(vars(dat)) -->

<!-- all_di <- all_di  %>% -->
<!--   mutate(skew_change = skew_percentile - skew_percentile_excl, -->
<!--          even_change = simpson_percentile - -->
<!--            simpson_percentile_excl) -->

<!-- ggplot(filter(all_di, s0 > 2), aes(x = log_nparts, y = skew_change)) + -->
<!--   geom_point() + -->
<!--   xlim(0,15) + -->
<!--   facet_wrap(vars(dat)) -->

<!-- ``` -->

<!-- ```{r hists} -->

<!-- ``` -->
